Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals

### 2022-04-16 draft_BLC/BLC_AM-ID_manuscript.tex
manuscript revisions, added literature

% !TeX document-id = {6d544fe2-4255-4108-8022-d809ff1d2c53}
% TeX program = xelatexmk %doesn't work, no xelatexmk option under 'build'. Changed default engine to xelatex (options-->configure-->build-->default build-->xelatex)
% see http://info.semprag.org/basics for a full description of this template
%\documentclass[times,linguex]{glossa}
\documentclass[12 pt]{article}
\usepackage[letterpaper]{geometry}
%\usepackage{times}
\geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}

@@

@@

%keep just the types here, and put the next 2 lines in the causes&effects section?
Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production}; 
(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}. %These compare productions during mixed-language use to participants' own productions while using a single language. The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness. 
The present study is concerned with the latter. In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use these interchangeably here to indicate any interaction between two sets of phonetic norms.

@@

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.

@@

@@

The vowel system of IE contains the monopthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}. %Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey). 
A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}. Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.

@@

Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf. Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf. transfer vs. interference \citep{grosjean2012attempt}; competence vs. performance interference \citep{paradis1993linguistic}). 
Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g. online processing costs \citep{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep{simonet2014phonetic}, context-awareness \citep{khattab2013phonetic}. What triggers this interaction? \cite{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
\begin{enumerate}
	\item The local point of switch itself
	\item Global co-activation of two languages -- \textsc{bilingual language mode} \citep{grosjean1998studying} 
\end{enumerate}

A number of studies have specifically manipulated language mode, both in the presence and absence of switching. Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep{olson2016role,tsui2019impact} or spontaneous conversation \citep{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer. (iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects of language mode, \cite{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.  Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \citep{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.

@@

Research comparing bilinguals speakers' phonological systems with monolinguals treats them as relatively stable over time. The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}. However, longitudinal studies %(citations) 
and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts. For example, \cite{bohn1992production} 
observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer. These have been interpreted as changes to long-term phonological representations.

Moreover, changes due to transfer are not necessarily unidirectional or irreversible. \cite{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language. This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range. A more recent study by \cite{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks). Based on a short-term longitudinal study, \cite{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2. For some sounds and features (though not others, cf. sec.\ref{asymmetry between sounds}), transfer was additive over time. How long these effects last in the absence of regular L2 input was not tested. 
% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)

The research discussed above concerns situations where the bilingual speaker is operating in any one of their languages. When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token. Thus, there are few direct measurements of the duration of short-term transfer effects. One study which measured this in a code-switching paradigm \citep{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized. Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
Note that all the studies discussed above measure VOT, which is a temporal feature. We have no apriori reason to assume that these durations generalize to vowel quality. However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}. Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode. We expect the intervening words between two subsequent targets to undo any residual effects of transfer.

@@

\subsubsection*{L1 vs. L2 status} 
Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another. Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories. In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf. \textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][]{guion2003vowel}.

In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep{tsui2019impact,elias2017effects, olson2013bilingual}, bidirectional convergence \citep{bullock2009trying, olson2016role}, divergence \citep{bullock2009trying,vsimavckova2018patterns}, and no influence \citep{muldner2019phonetics,schwartz2015language}. Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages. %(cf sec. \ref{sound systems}).

@@

Bengali has a four-way laryngeal contrast. Thus, VOT does not encode a binary voiced-voiceless distinction, and moreover is not the only relevant cue for the corresponding phonological categories \citep{dmitrieva2020acoustic}. This means that any effect of L1 Bengali on L2 English VOT is likely to be mediated by additional phonological factors. Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer.

@@

Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions. Studies that have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure. In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too. Once again, this emphasizes the importance of examining a wider range of sound contrasts.

@@

\end{enumerate}
\subsection{Acoustic analysis}\label{analysis}
The first iteration of each task was treated as a trial, and excluded from further analyses. An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens for analysis. Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel. This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.

@@

All statistical analyses were performed in R \citep{r}. The dependent variables were vowel height (F1) and backness (F2). We fitted a series of linear mixed effects models using the \emph{lme4} package \citep{lme4}, to examine the effect of experimental conditions on lobanov normalized F1 and F2 at each of the five points in the vowel (sec.\ref{analysis}). For all mixed models, the alpha criterion was set at $|t| > 2 $. The independent variables or fixed factors, along with their levels, are given in table \ref{table variables}. Subject and Item were treated as random factors.\\

@@

Our target in this analysis was to model vowel quality at different points in the vowel using the same mixed-effects model, allowing us to track if/how transfer changes over time. To do this, we arbitrarily chose a point in the vowel: F1 at 25\% (henceforth F1$\_$25) to identify an optimal model, and then fitted this to all other points.

Vowel quality and gender are expected to systematically affect F1. Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses: by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context. Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= .01854$), suggesting that the experimental condition affects vowel height. Following from our hypothesis \textbf{Asymmetry}, we expected the extent of change in vowel height between the conditions to be mediated by the identity of the vowel. Thus, we added a two-way interaction factor between Context and Vowel (Context*Vowel). As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= .0003837$). To assess whether the experimental paradigm affected F1, we added Task as a fixed effect. This marginally improved the model ($\chi^2(1) = 3.74, p= .05302$). Towards addressing our research question \textbf{Paradigm}, we wanted to assess whether the extent of shift differs across the two paradigms. However, no significant improvement was achieved by adding an interaction effect between Task and Context. Subsequent additions of other interactions, and removing each of the fixed effects in turn, did not lead to any further improvement.

Context and Vowel are both within-subject factors. Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \citep{brauer2018linear}. This improved the model fit ($\chi^2(5) = 11.43, p = .04342$). Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.

Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}). The following sections discuss the results for vowel height and backness respectively.

@@

Table \ref{table_f1_fixed_effects} summarizes the results of the fixed effects from the model for F1 at each point. As expected, we observe a significant effect of gender, such that male speakers have a lower mean F1 than female speakers. Vowel is not a significant predictor of F1. This is surprising, given that the phonological category is expected to systematically affect vowel quality. However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them. Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching. The lack of any independent effect of Context on F1 suggests that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels. Instead, the significant interaction between Context and Vowel indicates a lowering in \nt{2}, but not in \nt{\ae}, in the mixed-language condition. The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels, relative to the cued picture-naming task. This effect is strongest at the beginning of the vowel, and disappears by 45\%. This is an unexpected result, since we have no principled reason to expect code-switching to induce an across-board lowering of vowels. As noted earlier, there is no significant interaction of Task with any of the other fixed factors. We used a series of ANOVAs to compare the model elaborated above with a null model that differs in the absence of the main factor which concerns our hypothesis, Vowel*Context, at each point in the vowel. We find that the difference is not significant at 5\%, and increases over the course of articulating the vowel. \\
%ANOVA table add

>>>>

% !TeX document-id = {6d544fe2-4255-4108-8022-d809ff1d2c53}
% TeX program = xelatexmk %doesn't work, no xelatexmk option under 'build'. Changed default engine to xelatex (options-->configure-->build-->default build-->xelatex)
% see http://info.semprag.org/basics for a full description of this template
%\documentclass[times,linguex]{glossa}
\documentclass[12 pt]{article}
\usepackage[letterpaper]{geometry}
\usepackage{times}
\geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}

@@

\usepackage{easyReview}

@@

%keep just the types here, and put the next 2 lines in the causes&effects section?
Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production}; 
(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}. %These compare productions during mixed-language use to participants' own productions while using a single language. The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness. 
The present study is concerned with the latter. In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{transfer} here, to indicate any interaction between two sets of phonetic norms.

@@

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, \alert{but no difference in the extent of transfer between the two paradigms --and slightly different patterns of shift among the paradigms?}. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.

@@

Existing studies have studied phonetic transfer along different acoustic features. While discussing the literature in subsequent sections, we will often indicate the acoustic feature in parentheses following a citation, for clarity.

@@

The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}. %Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey). 
A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}. Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.

@@

Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf. Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf. transfer vs. interference \citep{grosjean2012attempt}; competence vs. performance interference \citep{paradis1993linguistic}). 
Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g. online processing costs \citep[][VOT]{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep[][vowel quality]{simonet2014phonetic}, context-awareness \citep[][phonological variables]{khattab2013phonetic}. What triggers this interaction? \cite{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
\begin{enumerate}
	\item The local point of switch itself
	\item Global co-activation of two languages -- \textsc{bilingual language mode} \citep{grosjean1998studying} 
\end{enumerate}

A number of studies have specifically manipulated language mode, both in the presence and absence of switching. Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer. (iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.  Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days (\cite{elias2017effects}; vowel quality), \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.

@@

Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time. The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}. However, longitudinal studies %(citations) 
and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts. For example, \cite[][vowel quality]{bohn1992production} 
observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer. These have been interpreted as changes to long-term phonological representations.

Moreover, changes due to transfer are not necessarily unidirectional or irreversible. \cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language. This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range. A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks). Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2. For some sounds and features (though not others, cf. sec.\ref{asymmetry between sounds}), transfer was additive over time. How long these effects last in the absence of regular L2 input was not tested. 
% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)

The research discussed above concerns situations where a speaker is operating in any one of their languages. When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token. Thus, there are few direct measurements of the duration of short-term transfer effects. One study which measured this in a code-switching paradigm \citep[][VOT]{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized. Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
Note that all the studies discussed above measure VOT, which is a temporal feature. We have no apriori reason to assume that these durations generalize to vowel quality. However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}. Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode. We expect the intervening words between two subsequent targets to undo any residual effects of transfer.

@@

\subsubsection*{L1 vs. L2 status} 
Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another. Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories. In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf. \textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.

In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}. Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages. %(cf sec. \ref{sound systems}).

@@

Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced-voiceless distinction. Moreover, VOT is not the only relevant cue for the corresponding phonological categories in the language \citep{dmitrieva2020acoustic}. This means that any effect of L1 Bengali on L2 English VOT is likely to be affected by additional phonological factors. Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer in the present study.

@@

Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions. Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure. In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too. Once again, this emphasizes the importance of examining a wider range of sound contrasts.

@@

\end{enumerate}
\subsection{Acoustic analysis}\label{analysis}
The first iteration of each task was treated as a trial, and excluded from further analyses. An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens (\nt{2}: 1137, \nt{\ae}: 1196) for analysis. Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel. This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel. Individual vocal tract differences affect formant frequencies across speakers. To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean (\alert{citation for lobanov normalization?}). Plotting the vowels (figure) shows a difference between male and female speakers, such that male speakers have lower formant frequencies. Normalizing by speaker does not get rid of this difference. Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each speaker's values as a scaled distance from the group mean). Statistical analyses were performed on these normalized values.

@@

All statistical analyses were performed in R \citep{r}. We used linear mixed effects models to test the effect of experimental conditions on vowel height (F1) and backness (F2) respectively. Since formants were extracted from five points over the course of articulating the vowel, time was coded numerically (1-5), and treated as an independent variable. This was scaled and centered prior to analysis and treated as a continuous predictor. Models were fit using the \emph{lme4} package \citep{lme4}, and \emph{lmerTest} \citep{lmerTest} was used to estimate p values for individual variables. For all mixed models, the alpha criterion was set at $|t| > 2 $. The dependent variables were Lobanov normalized F1 and F2, and we fitted separate models for these. The independent variables or fixed factors, along with their levels, are given in table \ref{table variables}. Subject and Item were treated as random factors.\\

@@

\subsection{Models}

\subsubsection*{Fixed effects}

Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs).

The experimental treatment of interest is Context, i.e. whether the target item was presented in a unilingual (English) or mixed (Bengali) context. Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance. We hypothesize that the direction of shift in the mixed condition is towards the corresponding L1 category. Thus, we expect the effect of Context on formant frequencies, i.e. the direction of shift, to be affected by the vowel category (c.f. the second hypothesis \bf{Asymmetry}). Thus, our main effect of interest is the interaction term Context*Vowel. This corresponds to the transfer pattern.

Our third hypothesis (\bf{Paradigm}) was that the language-mixing paradigm, or Task, will affect the extent and direction of transfer. If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.

Since we measured FFs at five points in the vowel, Time is expected to affect the FF. However, formant dynamics are expected to depend on the vowel category, and so we expect the interaction of Vowel*Time to predict formant frequency. If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.

\subsubsection*{Random effects structure}

Subject and Item were treated as random factors. Given our hypotheses, we included random intercepts for Subject and Item, by-subject random slopes for Vowel (formant targets for vowel categories differ across individuals) and Context (individuals differ in their response to the experimental condition), and a by-item random slope for Context (the effect of experimental conditions differ across words). \alert{(convergence problems with by-word random slope for context-- changing the optimizer to bobyqa solved the problem)}

We tested the significance of our hypothesized explanatory variables using t-values and p-values for individual variables in the models, and tested overall model fit by comparing the full model to a corresponding null model that lacks the variable of interest using ANOVAs. The following subsections summarize and discuss the model outputs for F1 and F2 respectively:

\subsection{Power analysis}
We were able to recruit 10 participants. To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the simr package \citep{simr} in R. The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).

Power is a function of sample size and effect size. Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}. This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$. Effect sizes in simr are expressed as unstandardized effect sizes (model estimates). In the absence (to the best of our knowledge) of literature on typical effect sizes for spectral differences in cross-language vowel production, we made a conservative estimate of effect sizes as |$\beta$| > 0.1. This amounts to the assumption that the smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}) is 0.1, and we want to confirm whether the study has enough power to detect such an effect, if it were to exist. Note that this is an arbitrary assumption; only more discussion among researchers in this field will lead to a consensus on what is considered a typical cross-language effect. 
In the  Calculations of d values from our models using the package EMAtools \citep{ematools} show that they range from 0.01 to 1.06 (no need).

While differences between paradigms have been discussed in previous literature and theoretically motivated, there is no work yet which has measured differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline. Since we are assuming $\beta$ = 0.1 to be the smallest difference across conditions that we are interested in, we will assume that a difference of similar magnitude across paradigms will be of theoretical interest to us. Thus, the power analysis fixes effect sizes of both variables at $\beta$ = 0.1, and report the power.

--> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit. While we have done our best to follow available literature, we would love to have more feedback on this. 
\subsubsection*{transfer term: vowel*context}
To calculate the power for the interaction term using simr, we set the estimate for the term to 0.05, and compare the full model against a model that lacks the interaction term. On the basis of 100 simulations, the power for

\subsubsection*{paradigm differences: vowel*context*task}

@@

\subsubsection*{Model output summary}

First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects. The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= -2.330e-01, standardized error (SE)=  5.244e-02, t= -4.44, p=0.0002). Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$). To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction. The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).

To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects. This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003). We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16). Comparing this model to null models that lack the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better. Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed. However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$). Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 13.36, p= 0.001$). This shows that the three-way interaction is the best fit for the data.
\alert{later, during interpreting part? --> Want to see: is the effect of task local (decreases with time)? Since the effect of just task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli} --
Because an independent effect of task on FFs is unexpected (c.f. ), we wanted to see if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).

Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time. This term has a marginally significant effect ($\beta$= -4.604e-02, SE=  2.692e-02, t= -1.71, p= 0.08), but does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 3.41, p= 0.18$).

Table summary:

\subsubsection*{Interpretation}
